---
title: "Differential expression analysis"
author: "Alicia Pliego"
date: "3/16/2022"
output: html_document
---

```{r setup, include=FALSE, echo=FALSE, message=FALSE}
library("DESeq2")
library(tximport)
library(plotly)
library(pheatmap)
library(org.Hs.eg.db)
library(DOSE)
library(pathview)
library(clusterProfiler)
library(AnnotationHub)
library(ensembldb)
library(tidyverse)
  library(EnhancedVolcano)
library(ComplexHeatmap)
```


# Differential expression analysis using DESeq2

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# txi import quant files by genes from salmon into DESEQ2

counts <- read.table('/Users/aliciapliego/Projects/iPSC/data/Count_QC-raw-count_v3.txt', header = TRUE)
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
counts_nondup <- subset(counts, !duplicated(subset(counts, select=c(gene_name))))
rownames(counts_nondup) <- counts_nondup$gene_name
counts_nondup$gene_name <- NULL
```


```{r,  echo=FALSE, warning=FALSE, message=FALSE}
metadata =readxl::read_excel('/Users/aliciapliego/Projects/iPSC/data/metadata_v2.xlsx')

fig <- plot_ly(
  type = 'table',
  header = list(
    values = c("<b>Samples</b>", names(metadata)),
  align = c('left', rep('center', ncol(metadata))),
  line = list(width = 1, color = 'black'),
  fill = list(color = 'rgb(235, 100, 230)'),
  font = list(family = "Arial", size = 14, color = "white")
  ),
  cells = list(
    values = rbind(
      rownames(metadata), 
      t(as.matrix(unname(metadata)))
    ),
    align = c('left', rep('center', ncol(metadata))),
    line = list(color = "black", width = 1),
    fill = list(color = c('rgb(235, 193, 238)', 'rgba(228, 222, 249, 0.65)')),
    font = list(family = "Arial", size = 12, color = c("black"))
  ))

fig
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}

dds <- DESeqDataSetFromMatrix(countData = round(counts_nondup),
                              colData = metadata,
                              design = ~ Type)
```

### Samples to compare in the Differential Analysis
Excluding sample tw047 from the analysis, a Differential expression analysis was performed:


```{r, echo=FALSE, warning=FALSE, message=FALSE}
#do filtering steps as for the clustering analysis
is_expressed <- assay(dds) >= 10
keep <- rowSums(assay(dds) >= 10) >= 2
dds <- dds[keep,]
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
dds <- DESeq(dds)
res <- results(dds)
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
resLFC <- lfcShrink(dds, coef= "Type_tw029_vs_av111", type="apeglm")
```

## Comparison of the different groups
Summary of differentially expressed genes:
```{r, echo=FALSE, warning=FALSE, message=FALSE}
resOrdered <- res[order(res$pvalue),]
summary(res)
```

Summary of differentially expressed genes.
Adjusting cut-off for and FDR of 0.05:
```{r, echo=FALSE, warning=FALSE, message=FALSE}
res05 <- results(dds, alpha=0.05, contrast = c("Type", "tw029", "av111"))
summary(res05)
```

Cutoff values:
  pvalue  = 10e-16
  FCcutoff = 1.5

### Volcano plot 
Genes upregulated in the treated sample are in the right of the volcano plot, whilst downregulated genes are in the left hand side.
```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.width= 8, fig.height = 9, fig.align='center'}
EnhancedVolcano(res05,
    lab = rownames(res05),
    title = 'tw029 vs av111',
    pCutoff = 10e-16,
    FCcutoff = 1.5,
    x = 'log2FoldChange',
    y = 'pvalue')

```

### Interactive volcano plot
```{r, echo=FALSE, warning=FALSE, message=FALSE}
res05_df <- as.data.frame(res05)
res05_df$log10P <- -log10(res05_df$pvalue) 
```

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.width= 8, fig.height = 9, fig.align='center'}
library(plotly)

vline <- function(x = 0, color = "red") {
  list(
    type = "line", 
    y0 = 0, 
    y1 = 1, 
    yref = "paper",
    x0 = x, 
    x1 = x, 
    line = list(color = color)
  )
}
hline <- function(y = 0, color = "blue") {
  list(
    type = "line", 
    x0 = 0, 
    x1 = 1, 
    xref = "paper",
    y0 = y, 
    y1 = y, 
    line = list(color = color)
  )
}


fig <- plot_ly(data = res05_df, x = ~res05_df$log2FoldChange, y = ~res05_df$log10P,
               text= rownames(res05_df), color = ~res05_df$log2FoldChange) %>%
  layout(xaxis = list(title = 'log2FoldChange'), font=t, plot_bgcolor = "#e5ecf6",
         yaxis = list(title = '-log10(Pval)'), legend = list(title=list(text='Legend Title')), shapes = list(vline(2.5), hline(20)))

fig 
```

### Counts of most differentially expressed gene

Expression of some microglia markers.
```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.width= 5, fig.height = 5, fig.align='center'}
# Save plotcounts to a data frame object
d <- plotCounts(dds, gene="TREM2", intgroup="Type", returnData=TRUE)

# Plotting the CTNNB1 normalized counts, using the samplenames (rownames of d as labels)
ggplot(d, aes(x = Type, y = count, color = Type)) + 
  geom_point(position=position_jitter(w = 0.1,h = 0)) +
  geom_text_repel(aes(label = rownames(d))) + 
  theme_bw() +
  ggtitle("TREM2") +
  theme(plot.title = element_text(hjust = 0.5))
```


```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.width= 5, fig.height = 5, fig.align='center'}
# Save plotcounts to a data frame object
par(mfrow=c(2,3))

d <- plotCounts(dds, gene="MPEG1", intgroup="Type", returnData=TRUE)

# Plotting the CTNNB1 normalized counts, using the samplenames (rownames of d as labels)
ggplot(d, aes(x = Type, y = count, color = Type)) + 
  geom_point(position=position_jitter(w = 0.1,h = 0)) +
  geom_text_repel(aes(label = rownames(d))) + 
  theme_bw() +
  ggtitle("MPEG1") +
  theme(plot.title = element_text(hjust = 0.5))


```

### Counts of other Differenatially expressed genes

From 9 out the 10 genes microglia signature from Mancuso 2019 does not seem that any gene is DE, however, in av111_001 shows different pattern of expression in some genes, which can be the reason of the difference with respect to the other groups.

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.width= 8, fig.height = 5, fig.align='center'}
# Save plotcounts to a data frame object

par(mfrow=c(3,3))

plotCounts(dds, gene="CX3CR1", intgroup="Type")
plotCounts(dds, gene="TMEM119", intgroup="Type")
plotCounts(dds, gene="P2RY12", intgroup="Type")
plotCounts(dds, gene="ELMO1", intgroup="Type")
plotCounts(dds, gene="P2RY13", intgroup="Type")
plotCounts(dds, gene="CSF1R", intgroup="Type")
plotCounts(dds, gene="SFMBT2", intgroup="Type")
plotCounts(dds, gene="IFNGR1", intgroup="Type")
plotCounts(dds, gene="GPR34", intgroup="Type")
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# Further filter DE genes based on Padj = 0.01,  and FC>1.5(L2FC = 0.58)
res_filtered_pval <- subset(res05, padj <0.01)
res_filtered_all <- subset(res_filtered_pval, log2FoldChange >0.58)
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# Rank genes and export for GSEA analysis
x <- res_filtered_pval
x$fcsign <- sign(x$log2FoldChange )
x$logP=-log10(x$pvalue)
x$metric= x$logP/x$fcsign
x$Gene <- rownames(x)
y<-x[,c("Gene", "metric")]
filtered <- na.omit(y)
write.table(filtered ,file="/Users/aliciapliego/Projects/iPSC/data/DE_genes.rnk",quote=F,sep="\t",row.names=F)
```

## Heatmap of top 50 DE genes

```{r, echo=FALSE, warning=FALSE, message=FALSE}
#Select TOP 20 DE genes by pvalue
res_df<- as.data.frame(res_filtered_pval)
resOrdered   <- res_df[order(res_df$pvalue),]
resOrdered50 <- resOrdered[1:50,]
top50genes   <- rownames(resOrdered50)
top50genes   <- as.data.frame(top50genes)
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
#Normalize results
dds <- estimateSizeFactors(dds)
normalized_counts <- counts(dds, normalized=T) %>% 
                     data.frame() %>%
                     rownames_to_column(var="gene") 
write.table(normalized_counts ,file="/Users/aliciapliego/Projects/iPSC/data/normalized_counts.txt",quote=F,sep="\t",row.names=F)
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}

## This will bring in a column of gene symbols
top50_counts<- merge(normalized_counts, top50genes, by.x="gene", by.y="top50genes")
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}

top50_counts2 <- top50_counts[,-1]
rownames(top50_counts2) <- top50_counts[,1]
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}

vst <- vst(dds, blind=FALSE)
vst <- assay(vst)
vst <- as.data.frame(vst)
vst_sig <- vst[rownames(vst) %in% rownames(top50_counts2),]
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}

heat <- t(scale(t(vst_sig)))
```


```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.width= 10, fig.height = 14, fig.align='center'}
Heatmap(heat, 
      row_names_gp = gpar(fontsize = 8),
    top_annotation = HeatmapAnnotation(foo = anno_block(gp = gpar(fill = 2:4),
        labels = c("tw029", "av111"), 
        labels_gp = gpar(col = "white", fontsize = 10))),
    column_km = 2)
```


## GSEA results
### General enrichment of DE genes in tw029 vs av111

Enriched pathways with a pval <0.05 for the different microglia markers.

Link to enrichment results:

#### Abud signature
Read more [here](abud.Gsea/ABUD_SIGNATURE.html) 

#### Gosselin signature
Read more [here](gosselin.Gsea/GOSSELIN_SIGNATURE.html) 

#### Hasselman signature
Read more [here](hasselman.Gsea/HASSELMAN.html) 

#### Haensler signature
Read more [here](haensler.Gsea/HAENSLER.html) 

#### Tissue microglia signature
Read more [here](tissue_microglia.Gsea/TISSUE_MICROGLIA.html) 

#### Autophagy signature
Read more [here](autophagy.Gsea/AUTOPHAGY.html) 

#### Activated microglia signature
Read more [here](activated_microglia.Gsea/ACTIVATED_MICROGLIA.html) 

